Article Title

### 2013-07-29 sections/Abstract.tex
Expanded Covariance Estimation Theory section.

Correlations between the spiking activity of pairs of neurons are among the most familiar and useful descriptive statistics of neural activity. Multineuronal recordings provide richer information than the equivalent number of neuron pairs. For example, the full covariance matrix reveals correlated activity across the entire population as well as partial correlations between pairs.  Estimation of covariance matrices from finite recordings can be improved by regularization, \emph{i.e.}\;by imposing some kind of structure.  Optimal regularization must be determined empirically since its performance depends on how well the imposed structure matches the underlying regularities. For example, we can use low-rank parameterizations of the covariance matrix to account for common activity fluctuations across the recorded population. Conversely, if correlations are strongly influenced by a small fraction of possible pairwise linear associations between the observed neurons, we can impose sparsity on the inverse covariance matrix. We can also use a `sparse+low rank’ inverse covariance representation to account for both common fluctuations and pairwise interactions.

>>>>

Correlations between the spiking activity of pairs of neurons are among the most familiar descriptive statistics of neural activity. Multineuronal recordings provide richer information than the equivalent number of neuron pairs. For example, the full covariance matrix reveals correlated activity across the entire population as well as partial correlations between pairs.  Estimation of covariance matrices can be improved by regularization, \emph{i.e.}\;by imposing some kind of structure.  Optimal regularization must be determined empirically since its effect depends on how closely the imposed structure matches the underlying regularities. For example, we can use low-rank parameterizations of the covariance matrix to account for common fluctuations across the recorded population. Conversely, if correlations are strongly influenced by a small fraction of pairwise linear associations between the observed neurons, we can impose sparsity on the inverse covariance matrix. We can also use a `sparse+low rank’ inverse covariance representation to account for both common fluctuations and pairwise interactions.

### 2013-07-29 sections/Theory.tex
Expanded Covariance Estimation Theory section.

Let $x_i \in \mathcal X, i=1\ldots,x_n$ denote a sample of consequtive observations of population activity of $p$ neurons in time bins $i$.  
The activity of each neuron is the real-valued firing rate, thus  $\mathcal X \in \mathbb R^{p\times 1}$.  
We assume that observations are drawn from some true cumulative probability distribution $F: \mathcal X \mapsto R$. Then let $\hat F_n$ denote the empirical distribution function obtained from the observed sample, which itself is a random variable.
\begin{equation}
\hat F_n(x) = \sum\limits_{i=1}^n \mathbf{1}(x > x_i)
\end{equation}
where $\mathbf 1(x_i \le x)$ is the indicator function which equals 1 when all elements of $x$ are greater than the corresponding elements of $x_i$ and 0 otherwise.

@@

The true covariance matrix $\Sigma \in \Theta$ is defined as a function of the true probability distribution
\begin{equation}
\Sigma(F) = \int\limits_{z\in\mathcal X} (z - \bar x)(z - \bar x)^\T \dif F(z)
\quad\mbox{where}\;
\bar x = \int\limits_{z\in\mathcal X} z \dif F(z)
\end{equation}
The \emph{plug-in estimator} is obtained by plugging the empirical distribution in place of the true distribution
\begin{equation}
\begin{split}
\Sigma(\hat F_n) = \int\limits_{z\in\mathcal X} (z - \bar x)(z - \bar x)^\T \dif \hat F_n(z) 
\equiv \frac  1 n\sum\limits_{i=1}^n (x_i-\bar x)(x_i - \bar x)
\\
\mbox{where}\;\bar x = 
\int\limits_{z\in \mathcal X}z \dif \hat F_n(z) \equiv \frac 1 n\sum\limits_{i=1}^n x_i
\end{split}
\end{equation}
The plugin estimator is known to be biased toward zero because the estimate of the mean $\bar x$ absorbs a fraction of the variance.

The usual unbiased estimate, the sample covariance matrix, rescales the plug-in estimator compensates for the bias
\begin{equation}
\hat\Sigma(\hat F_n) = \frac n {n - d} \Sigma(\hat F_n) \equiv \frac 1 {n-d} \sum\limits_{i=1}^n (x_i - \bar x) (x_i - \bar x)^\T
\end{equation}
When observations are independent, then $d=1$.  But when nearby samples are correlated, covariances become underestimated. Then we can use $d=cn$ where $c>1$ is called the \emph{overdispersion coefficient}, which may be estimated from data.

\subsubsection{Definition of regularization}
Now let's define a sequence of approximate biased estimates $\hat\Sigma_\gamma(\hat F_n)$ indexed by the \emph{regularization intensity}  $\gamma$ such that $\hat\Sigma_{\gamma=0}(\hat F_n)\equiv \hat\Sigma(\hat F_n)$, which become increasingly biased as $\gamma$ increases.

To evaluate the estimate, let's define the \emph{loss function} $\mathcal L : \Theta \times \Theta \mapsto \mathbb R$, which is minimized when its two arguments are equal.  Then \emph{excess loss} $\ell(\hat S, S) = \mathcal L(\hat S, S) - \mathcal L(S,S)$ assumes zero at the minimum.

The overall error, known as \emph{estimator risk} is expressed as
\begin{equation}
r_\gamma = \mathbb E_F\left[ \ell\left(\Sigma_\gamma(P_n),\Sigma\right) \right]
\end{equation}

The expected value of the approximate estimate 
$\Sigma_\gamma^{(n)} = \mathbb E_F\left[\hat \Sigma_\gamma(\hat F_n)\right]$ 
is biased with the amout of bias or \emph{approximaton error} expressed as  
\begin{equation}
b_\gamma^2 = \ell \left( \Sigma_\gamma^{(n)},\Sigma\right)
\end{equation}

Then the \emph{estimation error} 
\begin{equation}
\varepsilon_\gamma^2 = \mathbb E_F \left[ \ell\left(\hat \Sigma_\gamma(\hat F_n), \Sigma_\gamma^{(n)}\right) \right]
\end{equation}

When the loss function is the sum of squared error (Euclidean norm squared), which, for matricies, is called the Frobenius norm, $\mathcal L(\hat S,S) = \mathcal L_e(\hat S,S) = \|\hat S-S\|_F^2 = \Tr\left((\hat S-S)(\hat S-S)^\T\right)$, the estimation error is called \emph{variance} and the overall error is expressed as the the sum 
\begin{equation}
r_\gamma =  b_\gamma^2 + \varepsilon_\gamma^2
\end{equation}

With other loss functions, the overall error is also an increasing function of both kinds of errors, but not necessarily a simple sum.

One other popular loss function for covariance estimation arises from the theory of normal distributions.  Under the assumption of normality  with zero mean, the log likelihood of the covariance matrix $\Sigma$ given the independent sample $z_1,\ldots,z_m$, with $S = \tfrac 1 m \sum\limits_{i=1}^m z_i z_i^\T$,  
\begin{equation}
L\left(\Sigma \mid S\right) = -\tfrac m 2 \ln(2\pi) - \tfrac m 2 \ln \det \Sigma - \tfrac m 2 \Tr(\Sigma^{-1} S)
\end{equation}
Then the Gaussian loss function $\mathcal L_g$ is constructed by rescaling the normal log likelihood and dropping the constant term
\begin{equation}
\begin{split}
\mathcal L_g(\hat S,S) = &-\frac 2 m L\left(S \mid \hat S \right) - \ln(2\pi) 
\\ 
\equiv & \ln \det \hat S + \Tr(\hat S^{-1} S) 
\end{split}
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat S,S) = \mathcal L_g(\hat S,S) - \mathcal L_g(S,S) 
= -\ln \det \left(\hat S^{-1} S\right) + \Tr\left(\hat S^{-1}S\right) - p
\end{equation}
is known as \emph{entropy loss} \citep{James:1961}.

Despite the fact that entropy loss is inspired by the normal theory, the choice of a loss function is not equivalent to assuming a specific form of $F$. The loss function only measures the distance between the estimate and the true value of a parameter of the true distribution but not the distance between distributions themselves. Minimizing the loss function improves the estimate under any true distribution.

The task of regulazation comprises the choice of the \emph{regularization path} $\hat \Sigma_\gamma(\hat F_n)$ and the optimal location $\hat\gamma = \gamma(\hat F_n)$ on that path in oder to minimize the risk $r_\gamma$.  
By definition $\hat\Sigma_{\gamma=0}(\hat F_n)$ is an unbiased estimator with $b_{\gamma=0}^2 = 0$.

Both $\mathcal L_e(\hat\Sigma,\Sigma)$ and $\mathcal L_g(\hat\Sigma,\Sigma)$ are linear with respect to $\Sigma$ in the sense that
\begin{equation}
\mathcal L\left(\hat \Sigma,\alpha S_1 + (1-\alpha)S_2\right) 
\equiv 
\alpha\mathcal L(\hat\Sigma,S_1) + (1-\alpha)\mathcal L(\hat\Sigma,S_2)
\end{equation}
If we ensure that  $\hat\Sigma_Y$ is unbiased, i.e.\;$\mathbb E_Y\left[\hat \Sigma_Y\right]=\Sigma$, then this linearity implies that 
\begin{equation}
\mathbb E_Y \left[\mathcal L(\hat\Sigma,\hat\Sigma_Y)\right]
=
\mathcal L\left(\hat\Sigma,\mathbb E_Y \left[\hat\Sigma_Y\right]\right)
= 
\mathcal L(\hat\Sigma,\Sigma)
\end{equation}
In other words, the empirical loss $\mathcal L(\hat\Sigma,\hat\Sigma_Y)$ is an ubiased estimator of true loss $\mathcal L(\hat\Sigma,\Sigma)$ provided that $\hat\Sigma_Y$ is unbiased and $\hat\Sigma$ and $\hat\Sigma_Y$ are estimated from independent samples.  Note that $\hat\Sigma$ does not need to be unbiased.

In practice, recording a separate independent sample may not be sensible.  Instead, the recorded data a split into a training sample and validation sample.  One popular approach is $K$-fold \emph{cross-validation}, in which the data are split into $K$ roughly equal-sized subsets.  Each subset then successively serves as the validation  sample while the rest of the data make up the training sample. This procedure yields $k$ estimates of the loss, although not quite independent.

\subsection*{Covariance estimators}
It has long been recognized that the empirical covariance 
\begin{equation}
\hat S = \frac 1 n \sum\limits_{i=1}^n x_i x_i^\T
\end{equation}
is statistically inadmissible, meaning that larger sample sizes are required to reduce the loss function to the same level as other, more statistically efficient covariance estimators.

Covariance estimators can be made more efficient by two broad approaches, which are not mutually exclusive: (1) deliberately bias the estimator in order to reduce its variability and (b) reduce the dimensionality of the estimator by choosing a specific parametric model.

\subsubsection*{Shrinkage estimators}
The most familiar example of the first approach is the linear shrinkage estimator, calculated is a linear mixture of the empirical covariance estimate $\hat S$ and another low-variance target estimator $T$:
\begin{equation}
S_{shrink} = (1-\lambda) \hat S + \lambda T
\end{equation}
A popular target covariance is the identity matrix scaled by the mean sample variance  $T = \frac 1 p \Tr(\hat S) I$.
The optimal value of shrinkage estimate with respect to $\mathcal L_e$ can be estimated analytically from the data \citep{Ledoit:2004,Schafer:2005}.

\subsubsection*{Dimensionality reduction}
The second broad approach is to impose a structure on the covariance estimate by reducing the number of free parameters in one of its parameterizations.

>>>>

Let $x_i \in \mathcal X, i=1\ldots,x_n$ denote a sample of consequtive observations of population activity of $p$ neurons in time bins $i$.  
The activity of each neuron is the real-valued firing rate, thus  $\mathcal X = \mathbb R^{p\times 1}$.  
We assume that observations are drawn from some true cumulative probability distribution $F: \mathcal X \mapsto R$. Then let $\hat F_n$ denote the empirical distribution function obtained from the observed sample, which itself is a random variable.
\begin{equation}
\hat F_n(x) = \frac 1 n \sum\limits_{i=1}^n \mathbf{1}(x \ge x_i)
\end{equation}
where $\mathbf 1(x \ge x_i)$ is the indicator function which equals 1 when all elements of $x$ are greater than the corresponding elements of $x_i$ and 0 otherwise.

@@

The true covariance matrix $\Sigma \in \Theta$ is defined as a function of the true probability distribution
\begin{equation}
\Sigma = \int\limits_{z\in\mathcal X} (z - \bar x)(z - \bar x)^\T \dif F(z)
\quad\mbox{where}\;
\bar x = \int\limits_{z\in\mathcal X} z \dif F(z)
\end{equation}
The \emph{plug-in estimator} $\hat\Sigma$ is obtained by plugging the empirical distribution $F_n$ in place of the true distribution $F$
\begin{equation}
\begin{split}
\hat\Sigma = \int\limits_{z\in\mathcal X} (z - \bar x)(z - \bar x)^\T \dif \hat F_n(z) 
\equiv \frac  1 n\sum\limits_{i=1}^n (x_i-\bar x)(x_i - \bar x)
\\
\mbox{where}\;\bar x = 
\int\limits_{z\in \mathcal X}z \dif \hat F_n(z) \equiv \frac 1 n\sum\limits_{i=1}^n x_i
\end{split}
\end{equation}
The plug-in estimator is known to be biased toward zero because the estimate of the mean $\bar x$ absorbs a fraction of the variances.

The usual unbiased estimate, the sample covariance matrix, rescales the plug-in estimator to compensate for the bias
\begin{equation}
\hat\Sigma_0 = \frac n {n - d} \hat\Sigma \equiv \frac 1 {n-d} \sum\limits_{i=1}^n (x_i - \bar x) (x_i - \bar x)^\T
\end{equation}
When observations are independent, then $d=1$.  But when nearby samples are correlated, the estimate of the mean absorbs more of the variance, becomes \emph{overdispersed}. As a result the covariances become underestimated. Then we can use $d=cn$ where $c>1$ is called the \emph{overdispersion coefficient}, which may be estimated from the data.

\subsubsection{Bias-variance tradeoff}
Let $\hat\Sigma_\gamma$ be a family of approximate (biased) covariance estimates indexed by $\gamma$. To evaluate each estimate, let's define the \emph{loss function} $\mathcal L : \Theta \times \Theta \mapsto \mathbb R$ such that $\mathcal L(\hat S,S)$ is minimized when $\hat S = S$.  Then \emph{excess loss} $\ell(\hat S, S) = \mathcal L(\hat S, S) - \mathcal L(S,S)$ assumes zero at the minimum.

The overall error or \emph{estimator risk} 
\begin{equation}
r_\gamma = \mathbb E\left[ \ell\left(\hat\Sigma_\gamma,\Sigma\right) \right]
\end{equation}
can be decomposed into \emph{approximation error} or \emph{bias}   
\begin{equation}
b_\gamma^2 = \ell \left( \mathbb E\left[\hat \Sigma_\gamma\right],\Sigma\right)
\end{equation}
and \emph{estimation error} or \emph{variance}
\begin{equation}
\varepsilon_\gamma^2 = \mathbb E \left[ \ell\left(\hat \Sigma_\gamma, 
\mathbb E\left[\hat \Sigma_\gamma\right]\right) \right]
\end{equation}

When the loss function is the error sum of squares, which, for matrices, is called the Frobenius norm, 
\begin{equation}
\mathcal L_e(\hat S,S) = \|\hat S-S\|_F^2 = \Tr\left((\hat S-S)(\hat S-S)^\T\right)
\end{equation}
the bias-variance decomposition is the simple sum
\begin{equation}
r_\gamma =  b_\gamma^2 + \varepsilon_\gamma^2
\end{equation}

With other loss functions, the risk is also an increasing function of the two components, although not generally the sum.

Regularization is the game of minimizing risk $r_\gamma$ by reducing estimation error $\varepsilon_\gamma$ at the cost of increased bias $b_\gamma$, the \emph{bias-variance tradeoff}.  
We follow the convention that $\hat\Sigma_{\gamma=0}$ is the unbiased estimator $\hat\Sigma_0$, \emph{i.e.}\;$b_{\gamma=0}^2 = 0$.  Then as $\gamma$ increases, $\hat\Sigma_\gamma$ less variable but increasingly biased.  Thus $\hat\Sigma_\gamma$ defines the \emph{regularization path}, for which \emph{regularization intensity} $\gamma$ specifies the location along the path, which may be real-valued or discrete, or even a vector to steer the bias toward multiple targets.

\subsubsection{Evaluation of estimators}
Another popular loss  function for covariance estimation arises from the theory of normal distributions.  Under the assumption of normality  with zero mean, the log likelihood of the covariance matrix $\Sigma$ given the independent sample $z_1,\ldots,z_m$, with $S = \tfrac 1 m \sum\limits_{i=1}^m z_i z_i^\T$,  
\begin{equation}
L\left(\Sigma \mid S\right) = -\tfrac m 2 \ln(2\pi) - \tfrac m 2 \ln \det \Sigma - \tfrac m 2 \Tr(\Sigma^{-1} S)
\end{equation}
Then the Gaussian loss function $\mathcal L_g$ is constructed by rescaling the normal log likelihood and dropping the constant term
\begin{equation}
\begin{split}
\mathcal L_g(\hat S,S) = &-\frac 2 m L\left(S \mid \hat S \right) - \ln(2\pi) 
\\ 
\equiv & \ln \det \hat S + \Tr(\hat S^{-1} S) 
\end{split}
\end{equation}
The corresponding excess loss 
\begin{equation}
\ell_g(\hat S,S) = \mathcal L_g(\hat S,S) - \mathcal L_g(S,S) 
= -\ln \det \left(\hat S^{-1} S\right) + \Tr\left(\hat S^{-1}S\right) - p
\end{equation}
is known as \emph{entropy loss} \citep{James:1961}.

Despite the fact that entropy loss is derived from normal theory, the choice of a loss function is not equivalent to assuming a specific form of $F$. The loss function only measures the distance between the estimate and the true value of a parameter of the true distribution but not the distance between distributions themselves. Minimizing the loss function improves the estimate under any true distribution.

In practice, the true covariance matrix $\Sigma$ is not available and the estimator risk must be estimated from data. This requires a separate \emph{validation} empirical distribution $\hat F_m^\prime$ of $m$ observations sampled from $F$ independently of the \emph{training} distribution $\hat F_n$. The we can produce an independent unbiased covariance estimate $\hat \Sigma_0(\hat F_m^\prime)$.

Then the \emph{empirical risk} is 
\begin{equation}
\hat r_\gamma = \mathcal L\left(\hat\Sigma_\gamma,\hat\Sigma_0(\hat F_m^\prime)\right) 
\end{equation}

 The two loss functions $\mathcal L_e(\hat S,S)$ and $\mathcal L_g(\hat S,S)$ are particularly suitable for empirical risk estimation thanks to their linearity with respect to $S$ in the sense that 
\begin{equation}
\mathcal L\left(\hat S,\alpha S_1 + (1-\alpha)S_2\right) 
\equiv 
\alpha\mathcal L(\hat\Sigma,S_1) + (1-\alpha)\mathcal L(\hat S,S_2)
\end{equation}
which allows bringing the expectation inside 
\begin{equation}
\mathbb E\left[ \mathcal L\left(\hat\Sigma_r, \hat\Sigma_0(\hat F_m^\prime)\right) \right] 
=
\mathbb E\left[ \mathcal L\left(\hat\Sigma_\gamma, \mathbb E\left[\Sigma_0(\hat F_m^\prime)\right]\right) \right] 
=
\mathbb E\left[ \mathcal L\left(\hat\Sigma_\gamma, \Sigma\right) \right] 
\end{equation}
Then the expectation of the empirical risk is 
\begin{equation}
\mathbb E\left[\hat r_\gamma\right] 
= \mathbb E\left[\mathcal L(\hat\Sigma_\gamma,\Sigma)\right]
= r_\gamma + \mathcal L(\Sigma,\Sigma)
\end{equation}
This means, that our goal of minimizing the true loss $r_\gamma$ can be replaced with minimizing  the empirical loss $\hat r_\gamma$.

In practice, recording a separate independent validation sample may not be sensible.  Instead, the recorded data a split into a training sample and validation sample. In $K$-fold \emph{cross-validation} the data are split into $K$ roughly equal-sized subsets. Each subset then successively serves to construct the validation distribution $\hat F_m^\prime$  while the rest of the data is used to construct the training distribution $\hat F_n$.  This procedure produces $K$ estimates of the empirical risk, which are then averaged together to produce a better, averaged, estimate of the true risk.

Different regularization schemes can then be compared by comparing cross-validated empirical losses.

\subsubsection{Constructing a regularized estimator}
The task of designing a regularized estimators consists of defining a sequence of approximations $\hat\Sigma_\gamma$, called \emph{regularization path} and the optimal location on that path, \emph{regularization intensity} $\hat\gamma$ as  functions of the empirical distribution.

There are two basic approaches to design a family of approximations of a parameter that reduce the estimation error: \emph{dimensionality reduction} and \emph{shrinkage}.

Dimensionality reduction projects the estimate onto a low-dimensional manifold (subspace) within the $p(p+1)/2$-dimensional space of covariance matrices $\Theta$.  The variance in the projection is likely to be lower (if the manifold is smooth).  For example, factor analysis with $\nu$ factors projects (w.r.t.\;$\mathcal L_g$) $\hat\Sigma_0$ onto the low-dimensional manifold of covariance matrices representable by factor models. Thus factor analysis can be used as a regularized covariance estimator \citep{Fan:2008} with $\gamma$ controlling the number of factors (or, alternatively, the nuclear norm).  Another dimensionality reduction technique is \emph{covariance selection} \citep{Dempster:1972}, wherein a fraction of inverse covariance coefficients are constrained to zero and the closest (w.r.t.\;$\mathcal L_g$) solution to $\hat\Sigma_0$ is found.

Covariance shrinkage \citep{Schafer:2005,Ledoit:2004} uses a fixed, low-dimensional \emph{target estimate} $T$ and constructs the estimator as the linear mixture, with $\gamma \in [0,1]$,
\begin{equation}
\hat\Sigma_\gamma = \hat\Sigma_0  + \gamma (T-\hat\Sigma_0) 
\end{equation}
 The target can be a diagonal matrix with sample variances, an identity matrix scaled to the mean variance, a single factor model, or any other low-dimensional estimate.

An effective hybrid of dimensionality reduction and shrinkage can be obtained by using $L_1$-regularization techniques.  For example, graphical lasso \citep{Meinshausen:2006,Friedman:2008}, shrinks the inverse of the covariance toward zero as 
\begin{equation}
\hat\Sigma_\gamma = \arg\min\limits_S \mathcal L_g(S,\hat \Sigma_0) + \gamma \| S^{-1} \|_1
\end{equation}
This approach results in forcing many of the coefficients in the inverse of the estimate to zero, or ``sparsification'' of the the inverse.

Many regularization schemes combine multiple targets or multiple approaches, in which case $\gamma$ can become a vector of regularization parameters \citep{Schafer:2005,Fan:2011,Ma:2013} to be optimized in parallel.

\subsubsection{Optimization of the regularization parameter}
Analytic solutions when exist \citep{Ledoit:2004,Schafer:2005}, nested cross-validation otherwise.

\subsubsection{Interpretation of revealed structure}
Sparseness (dimensionality reduction) are key, but must be supported by data, not emph{ad hoc}. \citep{Fan:2006,Malmersjo:2013}.

\subsubsection{Out choices of regularized covariance estimates}
\begin{enumerate}[\qquad 1.\;\;]
\item sample covariance (benchmark)
\item covariance shrinkage 
\item factor model
\item graphical lasso
\item latent-variable graphical lasso
\end{enumerate}
